The document title

### 2021-03-16 manuscript/manuscript.Rmd
Add FS suggestions

Data collection took place at Porto Biomechanics Laboratory (LABIOMEP-UP) on an instrumented treadmill with built-in FP (AMTI Corporation, Watertown, USA). Before starting the data collection, participants were given time to warm up on the treadmill. The protocol consisted of walking and running in 13 incremental speeds, with 0º inclination, from 2 to 14 $\mathrm{km\cdot h^{-1}}$ with 1 $\mathrm{km\cdot h^{-1}}$ increment. Walking and running speeds ranged from 2 to 6 $\mathrm{km\cdot h^{-1}}$ and from 7 to 14 $\mathrm{km\cdot h^{-1}}$, respectively. Each walking speed lasted for 1 minute and each running speed for 30 seconds. To ensure single foot contact on each FP during the trial, feedback regarding location on the treadmill was provided by a researcher. Participants wore their own sports shoes throughout the trials.

@@

Activity monitors used were the GT9X Link model ($\mathrm{\pm16}$\textit{g} range; ActiGraph, Pensacola, USA) that incorporates primary and secondary triaxial accelerometers. As a manufacturer proprietary filter is applied on the primary, but not on the secondary accelerometer raw data, only data from the latter was used in this study, reducing filtering bias and enhancing the data processing replicability.

@@

Data from FP and accelerometers were collected at 1000 Hz and 100 Hz sampling frequency, respectively, and operated through the manufacturer supplied software (for the FP: Netforce, Version 3.5.1; AMTI Corporation, Watertown, USA; for the accelerometers: ActiLife version 6.13.3; Actigraph, Pensacola, USA). Raw GRF (expressed in N) and ACC (expressed in gravitational ACC units; 1\textit{g} = 9.807 $\mathrm{m\cdot s^{-2}}$) from the x, y and z vectors were exported to plain text files.

@@

Regression equations to predict pGFR and pLR for each accelerometer placement (ankle, lower back and hip) and vector (resultant and vertical) were developed by linear mixed models. Body mass (kg) was tested as a predictor in all models. Additionally, pACC and pAR were tested as predictors in the pGRF and pLR models, respectively. All predictors were tested as fixed effects and were shown to be significant. Subject and speed were tested as random effects, and the inclusion of both has showed to improve the models. Final models were chosen according to -2 log-likelihood statistics [@Field2012]. Traditional coefficient of determination ($\mathrm{R^2}$) was represented by conditional $\mathrm{R^2}$, that estimates the variance explained by the whole model [@Nakagawa2013].

@@

Models were validated using the leave-one-out cross-validation approach [@Staudenmayer2012]. For leave-one-out cross-validation each participant’s data were separated into a testing dataset (one participant at a time) with the remaining data in the training dataset. New linear mixed models with the same outcomes and predictors as determined for the entire sample were developed using the training dataset and then used to predict the outcome for the participant in the testing dataset. This process was repeated for all participants (`r max(sample_size$pGRF_N)` times). Data from the testing dataset was used in the remaining statistical analysis.

@@

To assess the models prediction accuracy, mean absolute error (MAE), mean absolute percent error (MAPE) and root mean square error (RMSE) were calculated. Also, methods agreement was evaluated with Bland-Altman plots by plotting the difference between the actual (FP measured) and predicted values against their mean. Bias was computed as the mean of these differences and the limits of agreement were obtained using \pm 1.96 standard deviation of the mean between actual and predicted values [@Bland1986]. In order to test whether there was a consistent over or underestimation of the acual values, one-sample \textit{t} tests were performed to assess if the difference between actual and predicted values are different than zero. Linear regressions were applied to identify if the magnitude of the mean between actual and predicted values influenced the magnitude of their difference. This was done to ascertain if the prediction accuracy was constant throughout the assayed magnitude range [@Giavarina2015]. These analyses were conducted separately for data from each accelerometer placement and outcome.

@@

```{r bmi_class}
BMI_avg <- sample_descriptives$BMI_mean
BMI_class <- NA
if (BMI_avg >= 18.5 & BMI_avg < 25) {
  BMI_class <- "normal weight"
} else if (BMI_avg >= 25 & BMI_avg < 30) {
  BMI_class <- "overweight"
} else if (BMI_avg >= 30 & BMI_avg < 35) {
  BMI_class <- "obesity class I"
} else if (BMI_avg >= 35 & BMI_avg < 40) {
  BMI_class <- "obesity class II"
} else if (BMI_avg >= 40) {
  BMI_class <- "obesity class III"
}
```

Our recruited sample presented an average BMI of `r sample_descriptives$BMI_mean` $\mathrm{kg\cdot m^{-2}}$, which is relatively high and classified as `r BMI_class`. 
It would be expected that participants with higher BMI values, due to low motor ability and physical capacity, are able to perform the trials only at the lowest speeds.
This is precisely what happened and, as can be observed in Figure 1, from 2 to 5 $\mathrm{km\cdot h^{-1}}$ the BMI median is around 30 $\mathrm{kg\cdot m^{-2}}$, decreasing to around 27 $\mathrm{kg\cdot m^{-2}}$ at 6 $\mathrm{km\cdot h^{-1}}$, which is the fastest walking speed, with a further reduction to less than 25 $\mathrm{kg\cdot m^{-2}}$ at the running speeds. Furthermore, only one participant with a BMI higher than 30 $\mathrm{kg\cdot m^{-2}}$ was able to run. This implies in a sample size reduction from `r filter(sample_size, acc_placement == "hip" & vector == "resultant" & speed == "2")$pGRF_N` at the slowest speed to `r filter(sample_size, acc_placement == "hip" & vector == "resultant" & speed == "14")$pGRF_N` at the fastest.

@@

As the main variables collected in this experiment are the peak magnitudes and rates of the GRF and ACC signals, there was a need to first establish their relationship. During locomotion, pACC increases linearly as pGRF increases. Such relation can be observed in the Figure 2, Panel A, for from hip-worn accelerometers and the resultant vector, and in Supplemental Figure S1 for all placements and the resultant vector and its vertical component. Moreover, for the same pACC values, the registered pGRF tended to be higher in subjects with higher BMI. It demonstrates that accelerometer data can adequately discriminate between different BMI categories. However, data from the pAR x pLR relationship show a greater dispersion, especially at higher values, and cannot discriminate so well the different BMI categories, as can be observed in Figure 2, Panel B, and Supplemental Figure S2.

@@

Afterwards, Bland-Altman plots were built to visualize the agreement between the actual and predicted pGRF and pLR, for both the resultant vector and its vertical component and all accelerometer placements. A good agreement in a Bland-Altman plot is observed when values tend to cluster around zero, showing no consistent over or underestimation of the actual values. Also, the majority of data points should lie within the limits of agreement, which are determined by \pm 1.96 standard deviation of the difference between actual and predicted values. Bland-Altman plots for the pGRF and pLR resultant vector from hip-worn accelerometers can be seen in Figure 3, and for all accelerometers placements and both resultant and vertical vectors in the Supplemental Figures S3 and S4. Values for all pGRF and pLR prediction equations developed tended to assemble around zero and between the limits of agreement. For the pGRF equations, a tendency to higher dispersion can be observed in the higher magnitudes, which correspond to the running trials, and especially in values derived from ankle-worn accelerometers (Supplemental Figure S3, Panels A and D). Prediction equations for pLR showed a more accentuated dispersion, and again in the higher magnitudes (Supplemental Figure S4). The pLR prediction model based on data from accelerometers placed at the ankle and the resultant vector displayed a bias different than zero ($p = `r myround(ba_bias_test_res_LR$ankle["p.value"], 2)`$). None of the other models presented a bias different than zero ($p > 0.05$), demonstrating that they do not consistently over or underestimate the pGRF and pLR values. Furthermore, linear regressions were applied to identify proportional bias, that is, to test whether pGRF and pLR prediction errors were systematically influenced by their magnitude. All equations based on ankle-worn accelerometer data, for both outcomes (pGRF and pLR) and vectors (resultant and vertical) showed a significant proportional bias ($p < 0.05$). Despite the statistical significance, these linear regressions presented a low $\mathrm{R^2}$ (highest $\mathrm{R^2} = `r myround(max(ba_regression_ankle_R2), 2)`$).

@@

Lastly, we compared the performance of our equations with a reference equation previously published. As the only comparable equation found was the one to predict pGRF vertical vector using hip-worn accelerometer data developed by Neugebauer et al [@Neugebauer2014], we applied it to our data and computed the MAE, MAPE and RMSE. Then, these indices were compared to the indices for our equation (Table 1). Apart from the overall equations performance, it was done a sub-analysis separating the walking and running trials. The results of these comparisons can be observed in the Table 2. Accuracy indices of our equations were all remarkably better compared to the reference equation, with a MAPE approximately 2 to 3 times lower. Furthermore, our equation performance was constant between walking and running, indicating that it can be used for both activities without loss of accuracy, which cannot be observed in the reference equation.

>>>>

Data collection took place at Porto Biomechanics Laboratory (LABIOMEP-UP) on an instrumented treadmill with built-in FP (AMTI Corporation, Watertown, USA). Before starting the data collection, participants were given time to warm up on the treadmill. The protocol consisted of walking and running in 13 incremental speeds, with 0º inclination, from 2 to 14 $\mathrm{km\cdot h^{-1}}$ with 1 $\mathrm{km\cdot h^{-1}}$ increment. Walking and running speeds ranged from 2 to 6 $\mathrm{km\cdot h^{-1}}$ and from 7 to 14 $\mathrm{km\cdot h^{-1}}$, respectively. Each walking speed lasted 1 minute and each running speed 30 seconds. To ensure single foot contact on each FP during the trial, feedback regarding location on the treadmill was provided by a researcher. Participants wore their own sports shoes throughout the trials.

@@

Activity monitors used were the GT9X Link model ($\mathrm{\pm16}g$ range; ActiGraph, Pensacola, USA) that incorporates primary and secondary triaxial accelerometers. As a manufacturer proprietary filter is applied on the primary, but not on the secondary accelerometer raw data, and therefore only data from the latter was used in this study, reducing filtering bias and enhancing the data processing replicability.

@@

Data from FP and accelerometers were collected at 1000 Hz and 100 Hz sampling frequency, respectively, and operated through the manufacturer supplied software (for the FP: Netforce, Version 3.5.1; AMTI Corporation, Watertown, USA; for the accelerometers: ActiLife version 6.13.3; Actigraph, Pensacola, USA). Raw GRF (expressed in N) and ACC (expressed in gravitational ACC units; 1$g$ = 9.807 $\mathrm{m\cdot s^{-2}}$) from the x, y and z vectors were exported to plain text files.

@@

Regression equations to predict pGFR and pLR in each vector (resultant and vertical) for each accelerometer placement (ankle, lower back and hip) were developed by linear mixed models. Body mass (kg) was tested as a predictor in all models. Additionally, pACC and pAR were tested as predictors in the pGRF and pLR models, respectively. All predictors were tested as fixed effects and were shown to be significant. Subject and speed were tested as random effects, and the inclusion of both has showed to improve the models. Final models were chosen according to -2 log-likelihood statistics [@Field2012]. Traditional coefficient of determination ($\mathrm{R^2}$) was represented by conditional $\mathrm{R^2}$, that estimates the variance explained by the whole model [@Nakagawa2013].

@@

Models were validated using the leave-one-out cross-validation approach [@Staudenmayer2012]. For this, each participant’s data were separated into a testing dataset (one participant at a time) with the remaining data in the training dataset. New linear mixed models with the same outcomes and predictors as determined for the entire sample were developed using the training dataset and then used to predict the outcome for the participant in the testing dataset. This process was repeated for all participants (`r max(sample_size$pGRF_N)` times). Data from the testing dataset was used in the remaining statistical analysis.

@@

To assess the models prediction accuracy, mean absolute error (MAE), mean absolute percent error (MAPE) and root mean square error (RMSE) were calculated. Also, methods agreement was evaluated with Bland-Altman plots by plotting the difference between the actual (FP measured) and predicted values against their mean. Bias was computed as the mean of these differences and the limits of agreement were obtained using \pm 1.96 standard deviation of the mean between actual and predicted values [@Bland1986]. In order to test whether there was a consistent over or underestimation of the acual values, one-sample \textit{t} tests were performed to assess if the difference between actual and predicted values are different than zero. Linear regressions were applied to identify if the magnitude of the mean between actual and predicted values influenced the magnitude of their difference. This was done to ascertain if the prediction accuracy was constant throughout the assayed magnitude range, which is known as proportional bias [@Giavarina2015]. These analyses were conducted separately for data from each accelerometer placement and outcome.

@@

Our recruited sample presented a broad range of body masses, from normal weight to severe obesity. It would be expected that participants with higher BMI values, due to low motor ability and physical capacity, are able to perform the trials only at the lowest speeds. This was precisely what happened and, as can be observed in Figure 1, from 2 to 5 $\mathrm{km\cdot h^{-1}}$ the BMI median was around 30 $\mathrm{kg\cdot m^{-2}}$, decreasing to around 27 $\mathrm{kg\cdot m^{-2}}$ at 6 $\mathrm{km\cdot h^{-1}}$, which was the fastest walking speed, with a further reduction to less than 25 $\mathrm{kg\cdot m^{-2}}$ at the running speeds. Furthermore, only one participant with a BMI higher than 30 $\mathrm{kg\cdot m^{-2}}$ was able to run. This implies in a sample size reduction from `r filter(sample_size, acc_placement == "hip" & vector == "resultant" & speed == "2")$pGRF_N` at the slowest speed to `r filter(sample_size, acc_placement == "hip" & vector == "resultant" & speed == "14")$pGRF_N` at the fastest.

@@

As the main variables collected in this experiment are the peak magnitudes and rates of the GRF and ACC signals, there was a need to first establish their relationship. During locomotion, pACC increases linearly as pGRF increases. Such relation can be observed in the Figure 2, Panel A, for from hip-worn accelerometers and the resultant vector, and in Supplemental Figure S1 for all placements and the resultant vector and its vertical component. Moreover, for the same pACC values, the registered pGRF tended to be higher in subjects with higher BMI. It demonstrates that accelerometer data can adequately discriminate between different BMI categories. However, data from the pAR x pLR relationship show a greater dispersion, especially at higher values, and the different BMI categories cannot be discriminated so well, as can be observed in Figure 2, Panel B, and Supplemental Figure S2.

@@

Afterwards, Bland-Altman plots were built to visualize the agreement between the actual and predicted pGRF and pLR, for both the resultant vector and its vertical component from all accelerometer placements. Bland-Altman plots for the pGRF and pLR resultant vector from hip-worn accelerometers can be seen in Figure 3, and for all accelerometer placements and both resultant and vertical vectors in the Supplemental Figures S3 and S4. Values for all pGRF and pLR prediction equations developed tended to assemble around zero and between the limits of agreement. For the pGRF equations, a tendency to higher dispersion can be observed in the higher magnitudes, which correspond to the running trials, and especially in values derived from ankle-worn accelerometers (Supplemental Figure S3, Panels A and D). Prediction equations for pLR showed a more accentuated dispersion, and again mainly in the higher magnitudes (Supplemental Figure S4). The pLR prediction model based on data from accelerometers placed at the ankle and the resultant vector displayed a bias different than zero ($p = `r myround(ba_bias_test_res_LR$ankle["p.value"], 2)`$). None of the other models presented a bias different than zero ($p > 0.05$), demonstrating that they do not consistently over or underestimate the pGRF and pLR values. All equations based on ankle-worn accelerometer data, for both outcomes (pGRF and pLR) and vectors (resultant and vertical) showed a significant proportional bias ($p < 0.05$). Despite the statistical significance, linear regressions presented a low $\mathrm{R^2}$ (highest $\mathrm{R^2} = `r myround(max(ba_regression_ankle_R2), 2)`$).

@@

Lastly, we compared the performance of our equations with a reference equation previously published. As the only comparable equation found was the one to predict pGRF vertical vector using hip-worn accelerometer data developed by Neugebauer et al [@Neugebauer2014], we applied it to our data and computed the MAE, MAPE and RMSE. Then, these indices were compared to the indices for our equation (Table 1). Apart from the overall equations performance, it was done a sub-analysis separating the walking and running trials. The results of these comparisons can be observed in the Table 2. Accuracy indices of our equations were all better compared to the reference equation, with a MAPE approximately 2 to 3 times lower. Furthermore, our equation performance was constant between walking and running, indicating that it can be used for both activities without loss of accuracy, which cannot be observed in the reference equation.
